Diagnostic Potential & the Interactive Dynamics of the Colorectal Cancer Virome

### 2017-10-21 doc/manuscript.Rmd
Updated document for submission.

---
title: Diagnostics & Interactive Dynamics of the Colorectal Cancer Virome
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology & Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family & Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
abstract: Viruses are assocaited with many human cancers, largely due to their mutagenic and functionally manipulative abilities. Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses. We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions while the virome role remains unknown. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
font: Times
fontsize: 12pt
---

@@

@@

## Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages (i.e. viruses that specifically infect bacteria) are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer. The gut virome (i.e. the virus community of the gut) therefore has the potential to impact health and disease. Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

@@

@@

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial OTU model to a model built using OVU relative abundances. The viral model performed as well as the bacterial model (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"])`), with the viral and bacterial models achieving mean area under the curve (AUC) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = asigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = asigfig)`, respectively **(Figure \ref{predmodel} A - B)**. To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data. The combined model yielded a modest but statistically significant performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**. The combined features from the virus and bacterial communities improved our ability to classify stool as belonging to individuals with cancerous colons.

@@

## Model for Virome & Cancer Progression
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a mechanistic link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in patients with colorectal cancer relative to those with healthy colons. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.

We have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

@@

## Conclusions

@@

## Materials and Methods
This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent. All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`. All associated source code is available at the following GitHub repository. Further detailed methods can be found in the supplemental methods.

https://github.com/SchlossLab/Hannigan_CRCVirome_Nature_2017

## Acknowledgments
The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. GD Hannigan was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528). PD Schloss was supported by funding from the National Institutes of Health (P30DK034933). MT Ruffin was supported by funding from the National Institutes of Health (5U01CA86400). The authors declare no competing interests.

## Figures

![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S rRNA gene sequence signatures, whole metagenomic signatures, and a combination of virome and 16S rRNA gene sequence signatures. A) An example ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon. B) Quantification of the AUC variation for each model, and how it compared to each of the other models based on 15 iterations. A pairwise Wilcoxon test with a false discovery rate multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01). C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S rRNA gene classification model when removed from the classification model. Mean is represented by a point, and bars represent standard error. D) Mean decrease in accuracy of each operational virus unit in the virome classification model. E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S rRNA gene and virome features.*\label{predmodel}](../figures/predmodel-viromebacteria.pdf){ width=100% }

![*Relative abundance correlations between bacterial OTUs and virome OVUs. A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OVUs (y-axis) with blue being positively correlated and red being negatively correlated. Bar plots indicate the viral (left) and bacterial (bottom) operational unit importance in their colorectal cancer classification models, such that the most important units are in the top left corner. B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OVUs. The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A). C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OVUs.*\label{correlations}](../figures/phage-bacteria-cor.pdf){ width=100% }

![*Community network analysis utilizing predicted interactions between bacteria and phage operational genomic units. A) Visualization of the community network for our colorectal cancer cohort. B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OVU. A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}](../figures/NetworkAndScatter.pdf){ width=80% }

![*Final model and working hypothesis from this study. A) Basic model illustrating the connections between the virome, bacterial communities, and colorectal cancer. B) Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}](../figures/WorkingHypothesis.pdf){ width=100% }

# Supplemental Materials

# Materials & Methods

@@

https://github.com/SchlossLab/Hannigan_CRCVirome_Nature_2017

@@

## Classification Modeling
Classification modeling was performed in R using the Caret package [@caretClassificatio:_U2Lit_1]. OTU, OVU, and OGU abundance data was preprocessed by removing features (OTUs, OVUs, and OGUs) that were present in less than thirty of the samples. This served both as an effective feature reduction technique and made the calculations computationally feasible. The binary random forest model was trained using the Area Under the receiver operating characteristic Curve (AUC) and the three-class random forest model was trained using the mean AUC. Both were validated using five-fold cross validation. Each training set was repeated five times, and the model was tuned for mtry values. For consistency and accurate comparison between feature groups (e.g., bacteria, viruses), the sample model parameters were used for each group. The maximum AUC during training was recorded across twenty iterations of each group model to test the significance of the differences between feature set performance. Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.

@@

@@

# Supplemental Figures

@@

![*Cohort and sample processing outline. Thirty subject stool samples were collected from healthy, adenoma (pre-cancer), and carcinoma (cancer) patients. Stool samples were split into two aliquots, the first of which was used for bacterial sequencing and the second which was used for virus sequencing. Bacterial sequencing was done using both 16S rRNA amplicon and whole metagenomic shotgun sequencing techniques. Virus samples were purified for viruses using filtration and a combination of chloroform (bacterial lysis) and DNase (exposed genomic DNA degradation). The resulting encapsulated virus DNA was sequenced using whole metagenomic shotgun sequencing.* \label{sampleproc}](../figures/sample-processing.pdf){ width=100% }

@@

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group. B) Sequence yield following quality control including quality score filtering and human decontamination.*\label{qualcontrol}](../figures/qualitycontrol.pdf){ width=100% }

@@

![*Length and coverage statistics. A) Heated scatter plot demonstrating the distribution of contig coverage (number of sequences mapping to each contig) and contig length for the virus metagenomic sample set. B) Scatter plot illustrating the distribution of operational viral unit (OVU) length and sequence coverage for the virus metagenomic sample set. C) Heated scatter plot demonstrating the distribution of contig coverage and length for the whole metagenomic sample set. D) Scatter plot illustrating the distribution of operational genomic unit (OGU) length and sequence coverage for the whole metagenomic sample set.*\label{contigqc}](../figures/qc-contiglengthcoverage.pdf){ width=75% }

@@

![*Operational genomic unit composition stats. A) Strip chart demonstrating the length and frequency of contigs within each operational genomic unit of the virome sample set. The y-axis is the operational genomic unit identifier, and x-axis is the length of each contig, and each dot represents a contig found within the specified operational genomic unit. B) Density plot (analogous to histogram) of the number of virome operational genomic units containing the specific number of contigs, as indicated by the x-axis. C-D) Sample plots as panels C and D, but for the whole metagenomic sample set.*\label{clustercontigqc}](../figures/contiglengthbycluster.png){ width=75% }

@@

![*Diversity calculations comparing cancer states of the colorectal virome, based on relative abundance of operational genomic units in each sample. A) NMDS ordination of community samples, colored for cancerous (green), pre-cancerous (red), and healthy (yellow). B) Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test. Comparisons (indicated on y-axis) in which the intervals cross the zero mean difference line (dashed line) were not significantly different. C) Shannon diversity and D) richness alpha diversity quantification comparing pre-cancerous (grey), cancerous (red), and healthy (tan) states.*\label{betaogu}](../figures/diversity-beta-ogu.pdf){ width=100% }

@@

![*Beta-diversity comparing disease states and the study negative controls. Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test. Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}](../figures/diversity-beta-ogu-negative.pdf){ width=100% }

@@

![*Comparison of bacterial 16S rRNA classification models with and without OTUs whose median relative abundance are greater than zero. A) Classification model performance (measured as area under the curve) for bacteria models using 16S rRNA data both with and without filtering of samples whose median was zero. Significance was calculated using a Wilcoxon rank sum test, and the resulting p-value is shown. The random area under the curve (0.5) is marked with a dashed line. B) Relative abundance of the six bacterial OTUs removed when filtered for OTUs with median relative abundance of zero. OTU relative abundance is seperated by healthy (red) and cancerous (grey) samples. Relative abundance of 1% is marked by the dashed line.*\label{16scompare}](../figures/filtered16S.pdf){ width=100% }

![*Transition of colorectal cancer importance through disease progression. A) Virus and B) 16S rRNA gene model performance (AUC) when discriminating all binary combinations of disease types. Blue line represents mean performance from multiple random iterations. C-E) Top ten important phage OVUs when classifying each combination of disease state, as measured by the mean decrease in accuracy metric. Mean is represented by a point, and bars represent standard error. Disease comparison is specified in the top left corner of each panel. F-H) Top ten important bacterial 16S rRNA gene OTUs for classifying each disease state combination.*\label{transitionmodels}](../figures/temporalmodels.pdf){ width=75% }

@@

![*ROC curves from A) virome and B) bacterial 16S three-class random forest models tuned on mean AUC. Each curve represents the ability of the specified class to be classified against the other two classes. C) Quantification of the mean AUC variation for each model based on 10 model iterations. A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that the models are significantly different (alpha = 0.01). D) Mean decrease in accuracy when virome operational genomic units and E) bacterial 16S OTUs are removed from the respective three-class classification models. Results based on 25 iterations. F) Relative abundance of the six most important virome OVUs in the model, with the most important on the right. Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=75% }

@@

![*Lysogenic phage relative abundance in disease states. Phage OVUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot. No disease groups were statistically significant.*\label{replicationstyles}](../figures/lysogenic-relabund.pdf){ width=50% }

>>>>

---
title: Diagnostic Potential & the Interactive Dynamics of the Colorectal Cancer Virome
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 12pt
mainfont: 'Helvetica'
---

&nbsp;

| ***Corresponding Author Information***
| Patrick D Schloss, PhD
| 1150 W Medical Center Dr. 1526 MSRB I
| Ann Arbor, Michigan 48109
| Phone: (734) 647-5801
| Email: pschloss@umich.edu

| **Article Class**:  Research Article

@@

# Abstract

Viruses are associated with many human cancers, largely due to their mutagenic and functionally manipulative abilities. Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses. We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions while the virome role remains unknown. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also predicted to be bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.

# Importance

Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its risk and severity have been linked to colonic bacterial community composition. Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities. We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences. The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community. These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages (i.e. viruses that specifically infect bacteria) are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer. The gut virome (i.e. the virus community of the gut) therefore has the potential to impact health and disease. Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

@@

# Results

@@

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial OTU model to a model built using OVU relative abundances. The viral model performed as well as the bacterial model (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"])`), with the viral and bacterial models achieving mean area under the curve (AUC) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = asigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = asigfig)`, respectively **(Figure \ref{predmodel} A - B)**. To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data. The combined model did not yield a statistically significant performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**.

@@

# Discussion
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a mechanistic link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in patients with colorectal cancer relative to those with healthy colons. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.

Using our findings here, we have begun to delineate the role the colonic virome plays in colorectal cancer in the form of a working hypothesis that will guide our future studies **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

@@

@@

# Materials and Methods

@@

https://github.com/SchlossLab/Hannigan_CRCVirome_mBio_2017

@@

## Classification Modeling
Classification modeling was performed in R using the Caret package [@caretClassificatio:_U2Lit_1]. OTU, OVU, and OGU abundance data was preprocessed by removing features (OTUs, OVUs, and OGUs) that were present in less than thirty of the samples. This served both as an effective feature reduction technique and made the calculations computationally feasible. The binary random forest model was trained using the Area Under the receiver operating characteristic Curve (AUC) and the three-class random forest model was trained using the mean AUC. Both were validated using five-fold nested cross validation to prevent over-fitting on the tuning paramters. Each training set was repeated five times, and the model was tuned for mtry values. For consistency and accurate comparison between feature groups (e.g., bacteria, viruses), the sample model parameters were used for each group. The maximum AUC during training was recorded across twenty iterations of each group model to test the significance of the differences between feature set performance. Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.

@@

# Funding Information
GD Hannigan was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528). PD Schloss was supported by funding from the National Institutes of Health (P30DK034933). MT Ruffin was supported by funding from the National Institutes of Health (5U01CA86400). The authors declare no competing interests.

# Acknowledgments
The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.

@@

# Figure Legends

![*Cohort and sample processing outline. Thirty subject stool samples were collected from healthy, adenoma (pre-cancer), and carcinoma (cancer) patients. Stool samples were split into two aliquots, the first of which was used for bacterial sequencing and the second which was used for virus sequencing. Bacterial sequencing was done using both 16S rRNA amplicon and whole metagenomic shotgun sequencing techniques. Virus samples were purified for viruses using filtration and a combination of chloroform (bacterial lysis) and DNase (exposed genomic DNA degradation). The resulting encapsulated virus DNA was sequenced using whole metagenomic shotgun sequencing.* \label{sampleproc}]()

![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S rRNA gene sequence signatures, whole metagenomic signatures, and a combination of virome and 16S rRNA gene sequence signatures. A) An example ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon. B) Quantification of the AUC variation for each model, and how it compared to each of the other models based on 15 iterations. A pairwise Wilcoxon test with a false discovery rate multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01). C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S rRNA gene classification model when removed from the classification model. Mean is represented by a point, and bars represent standard error. D) Mean decrease in accuracy of each operational virus unit in the virome classification model. E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S rRNA gene and virome features.*\label{predmodel}]()

![*Relative abundance correlations between bacterial OTUs and virome OVUs. A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OVUs (y-axis) with blue being positively correlated and red being negatively correlated. Bar plots indicate the viral (left) and bacterial (bottom) operational unit importance in their colorectal cancer classification models, such that the most important units are in the top left corner. B) Magnification of the boxed region in panel (A), highlighting the correlation between the most important bacterial OTUs and virome OVUs. The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as panel (A). C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OVUs.*\label{correlations}]()

![*Lysogenic phage relative abundance in disease states. Phage OVUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot. No disease groups were statistically significant.*\label{replicationstyles}](){ width=50% }

![*Final model and working hypothesis from this study. A) Basic model illustrating the connections between the virome, bacterial communities, and colorectal cancer. B) Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}]()

# Supplemental Figure Legends

@@

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group. B) Sequence yield following quality control including quality score filtering and human decontamination.*\label{qualcontrol}]()

@@

![*Length and coverage statistics. A) Heated scatter plot demonstrating the distribution of contig coverage (number of sequences mapping to each contig) and contig length for the virus metagenomic sample set. B) Scatter plot illustrating the distribution of operational viral unit (OVU) length and sequence coverage for the virus metagenomic sample set. C) Heated scatter plot demonstrating the distribution of contig coverage and length for the whole metagenomic sample set. D) Scatter plot illustrating the distribution of operational genomic unit (OGU) length and sequence coverage for the whole metagenomic sample set.*\label{contigqc}]()

@@

![*Operational genomic unit composition stats. A) Strip chart demonstrating the length and frequency of contigs within each operational genomic unit of the virome sample set. The y-axis is the operational genomic unit identifier, and x-axis is the length of each contig, and each dot represents a contig found within the specified operational genomic unit. B) Density plot (analogous to histogram) of the number of virome operational genomic units containing the specific number of contigs, as indicated by the x-axis. C-D) Sample plots as panels C and D, but for the whole metagenomic sample set.*\label{clustercontigqc}]()

@@

![*Diversity calculations comparing cancer states of the colorectal virome, based on relative abundance of operational genomic units in each sample. A) NMDS ordination of community samples, colored for cancerous (green), pre-cancerous (red), and healthy (yellow). B) Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test. Comparisons (indicated on y-axis) in which the intervals cross the zero mean difference line (dashed line) were not significantly different. C) Shannon diversity and D) richness alpha diversity quantification comparing pre-cancerous (grey), cancerous (red), and healthy (tan) states.*\label{betaogu}]()

@@

![*Beta-diversity comparing disease states and the study negative controls. Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test. Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}]()

@@

![*Comparison of bacterial 16S rRNA classification models with and without OTUs whose median relative abundance are greater than zero. A) Classification model performance (measured as area under the curve) for bacteria models using 16S rRNA data both with and without filtering of samples whose median was zero. Significance was calculated using a Wilcoxon rank sum test, and the resulting p-value is shown. The random area under the curve (0.5) is marked with a dashed line. B) Relative abundance of the six bacterial OTUs removed when filtered for OTUs with median relative abundance of zero. OTU relative abundance is separated by healthy (red) and cancerous (grey) samples. Relative abundance of 1% is marked by the dashed line.*\label{16scompare}]()

@@

![*Transition of colorectal cancer importance through disease progression. A) Virus and B) 16S rRNA gene model performance (AUC) when discriminating all binary combinations of disease types. Blue line represents mean performance from multiple random iterations. C-E) Top ten important phage OVUs when classifying each combination of disease state, as measured by the mean decrease in accuracy metric. Mean is represented by a point, and bars represent standard error. Disease comparison is specified in the top left corner of each panel. F-H) Top ten important bacterial 16S rRNA gene OTUs for classifying each disease state combination.*\label{transitionmodels}]()

@@

![*ROC curves from A) virome and B) bacterial 16S three-class random forest models tuned on mean AUC. Each curve represents the ability of the specified class to be classified against the other two classes. C) Quantification of the mean AUC variation for each model based on 10 model iterations. A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that the models are significantly different (alpha = 0.01). D) Mean decrease in accuracy when virome operational genomic units and E) bacterial 16S OTUs are removed from the respective three-class classification models. Results based on 25 iterations. F) Relative abundance of the six most important virome OVUs in the model, with the most important on the right. Line indicates abundance mean.*\label{threewaymodel}](){}

@@

![*Community network analysis utilizing predicted interactions between bacteria and phage operational genomic units. A) Visualization of the community network for our colorectal cancer cohort. B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OVU. A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}]()
